A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-01-25 analysis/paper/paper.Rmd
add a reference and highlights

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationship between entities. Different from traditional static networks, statistical approaches to network analysis, such as exponential random graph models (ERGMs), provides a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  These are the highlights. 
---

@@

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show evidence of imported prestige goods, such as ornaments and porcelains, in burials and some accumulation patterns in residential areas around the same time as European contact in the 17th century [@Chen2007; @LWandBM2020ornament]. The use of imported goods is also mentioned in historical records from the Spanish, indicative of the pursuit of prestige or wealth in those Indigenous societies [@Borao2009; @LiandWu2006]. For example, the document written by a Spanish Priest described that the Spanish soldier used carnelian beads to exchange natural resources with local Indigenous people since their high values in Indigenous culture [@LiandWu2006]. In addition, an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status by their community [@LiandWu2006]. Despite we observed the uneven distribution of prestige goods in burials, the degree of differentiation across burials over time remain unclear. The introduction of prestige goods to local Indigenous communities might occur earlier before a European presence became established, but was amplified during the European colonial period [@Borao2009; @LiandWu2006; @LWandBM2020ornament]. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals, and unequal access to trade networks [@Brumfiel1994; @Clark1994]. The observed uneven distribution may be explained as a result of social change from less social inequality to more social inequality when Indigenous societies in Northeastern Taiwan were involved in the complex trade network stimulated by the Europeans.

>>>>

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationship between entities. Different from traditional static networks, statistical approaches to network analysis, such as exponential random graph models (ERGMs), provides a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on Exponential random graph models (ERGMs) to model archaeological networks.
  2. We test social processes by simulating two sets of burial data from different time.
  3. Differences in network statistics implies changes in social structures influenced by foreign presence. 
  4. Bayesian ERGMs is efficient and useful for evaluating and interpreting networks. 
---

@@

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show evidence of imported prestige goods, such as ornaments and porcelains, in burials and some accumulation patterns in residential areas around the same time as European contact in the 17th century [@Chen2007; @LWandBM2020ornament]. The use of imported goods is also mentioned in historical records from the Spanish, indicative of the pursuit of prestige or wealth in those Indigenous societies [@Borao2009; @LiandWu2006]. For example, the document written by a Spanish Priest described that the Spanish soldier used carnelian beads to exchange natural resources with local Indigenous people since their high values in Indigenous culture [@LiandWu2006]. In addition, an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status by their community [@LiandWu2006]. Despite we observed the uneven distribution of prestige goods in burials, the degree of differentiation across burials over time remain unclear. The use of trade items as prestige goods in local Indigenous communities might occur earlier before a European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016]. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals, and unequal access to trade networks [@Brumfiel1994; @Clark1994]. The observed uneven distribution may be explained as a result of social change from less social inequality to more social inequality when Indigenous societies in Northeastern Taiwan were involved in the complex trade network stimulated by the Europeans.
